A lottery model of density-dependent selection in evolutionary genetics

### 2017-05-06 manuscript.tex
latest

Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition. This broadly follows the original spirit of MacArthur's $r/K$ scheme. More specifically, our aim is to begin merging some ecological realism to population genetics' time-dependent, genetically-focused view of evolution. We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.

@@

We then explore some time-dependent behavior of our extended lottery model. Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms in the presence of cyclical population density.

@@

We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work. This is an excellent approximation when the $n_i$ are all large. How ever, we also do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance. When considering new mutations, we therefore restrict our attention to begin with the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{lottery.pdf}
\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements. First, propagules are produced by adults which are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown). Lottery competition then occurs in each unoccupied territory (competition in only one territory is illustrated): each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory. Then occupied territories are freed up by adult mortality. In Eq. \eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').}
\end{figure}

@@

For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
This seems reasonable in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).

@@

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate role of selection in shaping the corresponding ``primary strategies''. To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.

Ideal environments are characterized by the near-absence of stress and disturbance, so that survival is easy $d\ll 1$ and growth and reproduction can be rapid $b\gg d$. The propagule density is determined by the magnitude of $b/d$, and in this case $L\gg 1$ (Appendix C).

@@

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$; it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}. In our model, this implies $L\ll 1$.

@@

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}. Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}. Only a tiny proportion of neutral and deleterious mutations fix (probability of order $1/N$ for neutrals).

Accounting for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article. We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects. Specifically, we consider the effect of mutations that change $b$, $c$, or $d$ by a fixed proportion $b_j=(1+\epsilon) b_i$, $c_j=(1+\epsilon) c_i$ or $d_j=(1-\epsilon) d_i$, where $j$ is is a novel mutant (Appendix C). Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits .

Not surprisingly, mutations which increase $c$ are effectively neutral when $L\ll 1$ and adaptive when $L\gg 1$. However, beyond this, selection in our density-dependent lottery model does not favor one trait improvement over another (Fig. 3). This applies regardless of the initial magnitude of $b$ or $d$. More specifically, in ideal and stressed environments, where Eq. \eqref{eq:master} applies, $b$ and $d$ mutations of size $\epsilon$ have identical fitness effects (Appendix C), and the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$.

In disturbed environments, $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on $d$, suggesting the primacy of $b$ under disturbance. However, this effect is easily swamped by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality. In Appendix C, we argue that $b$ and $d$ mutations should have comparable effects on fitness (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves and so $d$ cannot run ahead of $b$ for long (diminishing returns).

@@

In the previous section we only considered how $b$, $c$ and $d$ should respond in Grime's environmental extremes, based on invasion fitness. Here we further explore the low frequency behavior of Eq. \eqref{eq:master} to determine which types can coexist in a constant environment, and then consider the full time-dependent behaviour of Eq. \eqref{eq:master} in a cyclical environment.

In a constant environment, stable coexistence is possible in our extended lottery model. A $b$ specialist $i$ and $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$) can co-exist because then propagule density $L$ is frequency dependent, and so is the importance of competitive ability (Appendix D). This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit. A similar situation holds for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}. These forms of co-existence require density dependence (being mediated by $L$), and are not present in the classic lottery model. Coexistence is not possible between $b$- and $d$-specialists (Appendix D).

@@

Fig. \ref{fig:fluctuatingselection}a-c shows the behavior of Eq. \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth). Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating. Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$). Neither type has an advantage over a full environmental cycle, and they stably coexist. This is due to a combination of the storage and bounded density effects (recall that stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).

The classic lottery model (Eq. 1) gives dramatically different co-existence predictions because expansion to carrying capacity occurs immediately at the start of the summer (Fig. \ref{fig:fluctuatingselection}d-f). As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than required when we properly account for the growth in the abundance of each type using Eq. \eqref{eq:master} (keeping the other parameters the same). Previous models of the promotion of genetic variation via the storage effect  \citep{ellner_1994} similarly assume that the total number of offspring per iteration is constant, and would produce a similar error.

@@

\begin{figure}
\centering
\includegraphics[scale=0.7]{fluctuatingselection.pdf}
\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment requires a much greater $b$ advantage in the classic lottery model compared to our density-dependent extension of it when population density is seasonally cyclical. (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer). The $b$ specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$ specialist ($b=0.217$, $d=0.1$) (blue). (b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist does so at a lower rate. Total height (blue+black) is population density $N/T$. (c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist. (d-f) Same as (a-c) for the classic lottery model; the types no longer coexist. (g-i) Same as (d-f) where now $b = 0.0421$ for the $d$ specialist and the types coexist. For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed dispersal.} 
\end{figure}

@@

It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme, where $r=b-d$ is the maximal, low-density growth rate \citep{pianka_1972}. Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}. Contrary to predictions of an $r$/$K$ trade-off, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, this positive correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.

@@

$K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements. To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density per territory $L$. The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different genotypes. We leave these complications for future work.

Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations which can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}. Moreover, our formulation is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}). However, section ``Primary strategies and Grime's triangle'' suggests that, apart from the obvious irrelevance of territorial contests at low density, selection alone has no preference for fecundity, competitive ability or longevity. This could mean that one or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation. A separate model of trade-offs and mutation effects would be needed to explore this possibility.

Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$ specialists), may also be strongly $d$-selected, consistent with our findings. An effective way to reduce $d$ in the face of unavoidable physical destruction is to shorten the time to reproductive maturity --- short life cycles are a characteristically ruderal trait. Moreover, a recent hierarchical cluster analysis of coral traits did find a distinct ``ruderal'' cluster, but high fecundity was not its distinguishing feature. Rather, ruderals used brood- (as opposed to broadcast-) spawning, which could plausibly be a mechanism for improving propagule survivorship in disturbed environments \citep{darling_2012}.

One potential limitation of our model as a general-purpose model of density-dependent selection is its restriction to interference competition between juveniles for durable resources (lottery recruitment to adulthood), analogous to the ubiquitous assumption of viability selection in population genetics \citep[p. 45]{ewens_2004}. In some respects this is the complement of consumable resource competition models, which restrict their attention to indirect exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that consumable resources are spatially localized (e.g. due to restricted movement through soils), resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition). The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar genotypes that use identical resources rather than the stable co-existence of widely differing species with different resource preferences \citep{ram_2016}. We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources. More work is needed connecting resource competition models to the density-dependent selection literature, since most of the former has to date been focused on narrower issues of the role of competition at low resource availability \citep{aerts_1999,davis_1998,tilman_2007}.

@@

In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference. It seems that Eq. \eqref{eq:master} or something similar (and hopefully more analytically tractable) is unavoidable for the analysis of time-course genetic data because, fundamentally, selective births and deaths affect both abundances and frequencies, not one or the other in isolation. Moreover, some aspects of allele frequency change are intrinsically density-dependent. In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model''). This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist. This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}. It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}). In cyclical environments, polymorphisms can be stabilized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}. We leave the details of how our model might be applied in inference, including the crucial issue of its genetic drift predictions (providing a null model for neutral sites), for future work.

@@

In our representation of Grime's ideal environment, $d\ll 1$, $b_i/d_i \gg 1$ and $L\gg 1$, and so Eq. \eqref{eq:invad} becomes 
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j}{c_j/c_i + L-1}\frac{c_j}{c_i}-d_j\approx \frac{b_j}{L}\frac{c_j}{c_i}-d_j, \label{eq:highLinvad}
\end{equation}
where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i$ (ignoring mutations with very large effect sizes; incidentally, Eq. \eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this assumption is unrealistic for a rare mutant). Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium. Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy. The same argument applies for the high-$L$ interpretation of the stress regime.

@@

The effects of improving $b$ and $d$ are now no longer identical for given values of $d_{ie}$ and $d_{ig}$. A $b-$mutation $b_j=(1+\epsilon)b_i$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i \approx \epsilon$, whereas a $d$-mutation $d_{jg}=(1-\epsilon) d_{ig}$ gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$. Since overall mortality is high $d_i\approx 1$, $d_{g}$ can be small compared to $d_{e}$ and yet still make a greater contribution to invasion fitness than $b$. The case where $d_{ig}$ is so small that $d_{ig}\ll 1-d_i$ (i.e. $d_{ig}$ is an order of magnitude smaller than $1-d_i$ which itself an order of magnitude smaller than $d_{ie}$) effectively represents a hard constraint that $d$ cannot evolve. We do not deny this possibility, but this bring us back to the issue of mutations affect traits (the first two factors controlling the direction of evolution in section ``Primary strategies and Grime's triangle''), which we do not address in this manuscript.

>>>>

Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition. This broadly follows the original spirit of MacArthur's $r/K$ scheme. More specifically, our aim is to begin merging some ecological realism into population genetics' time-dependent, genetically-focused view of evolution. We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.

@@

We then explore some time-dependent behavior of our extended lottery model. Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms when population density is cyclical.

@@

We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work. This is an excellent approximation when the $n_i$ are all large. However, we also do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance. When considering new mutations, we therefore restrict our attention to begin with the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).

\begin{figure}
\centering
\includegraphics[scale=0.8]{lottery.pdf}
\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements. First, propagules are produced by adults and are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown). Lottery competition then occurs in each unoccupied territory (competition in only one territory is illustrated): each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory. Then occupied territories are freed up by adult mortality. In Eq. \eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').}
\end{figure}

@@

For the majority of this manuscript we assume that mortality only occurs in adults (Fig 2; setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}
This seems reasonable in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles.

@@

Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the role of selection in shaping the corresponding ``primary strategies''. To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.

Ideal environments are characterized by the near-absence of stress and disturbance, so that survival is easy ($d\ll 1$) and growth and reproduction can be rapid ($b\gg d$). The propagule density is determined by the magnitude of $b/d$, and hence $L\gg 1$ (Appendix C).

@@

Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein). Severe stress inhibits growth and reproduction, so that $b\ll 1$. In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate ($b/d\approx 1$); it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}. In our model, this implies $L\ll 1$.

@@

The expected direction of evolution is determined by three factors. First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to physiological trade-offs between them. Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}. Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically rather than be lost to drift is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.

Accounting for the first two factors is beyond the scope of this article. We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with comparable distributions of beneficial fitness effects. Specifically, we consider the effect of mutations that change $b$, $c$, or $d$ by a fixed proportion $b_j=(1+\epsilon) b_i$, $c_j=(1+\epsilon) c_i$ or $d_j=(1-\epsilon) d_i$, where $j$ is is a novel mutant (Appendix C). Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits.

Not surprisingly, mutations which increase $c$ are effectively neutral when $L\ll 1$ and adaptive when $L\gg 1$. However, selection in our density-dependent lottery model does not favor improvement in $b$ versus $d$. The same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$ in ideal and stressed (HD) environments.

In disturbed environments, where $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$, the argument for equal selection on $b$ and $d$ is more subtle and contingent. The bulk of the high mortality $d\approx 1$ is environmental and unavoidable. This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on total $d$, suggesting the primacy of $b$ under disturbance. However, this effect is easily matched by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality. In Appendix C, we argue that $b$ and $d$ mutations should contrive to have comparable effects on fitness in the disturbance case (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves.

@@

In the previous section we only considered how $b$, $c$ and $d$ should respond to selection in Grime's environmental extremes, based on invasion fitness. Here we further explore the low frequency behavior of Eq. \eqref{eq:master} to determine which types can coexist in a constant environment, and then consider the full time-dependent behaviour of Eq. \eqref{eq:master} in a cyclical environment.

In a constant environment, stable coexistence is possible in our extended lottery model. A $b$-specialist $i$ and $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$) can co-exist because then propagule density $L$ is frequency-dependent, and so is the importance of competitive ability (Appendix D). This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}; the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit. A similar situation holds for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}. These forms of co-existence require density dependence (being mediated by $L$), and are not present in the classic lottery model. Coexistence is not possible between $b$- and $d$-specialists in a constant environment (Appendix D).

@@

Fig.~\ref{fig:fluctuatingselection}a-c shows the behavior of Eq. \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth). Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating. Two types are present, a $b$-specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$-specialist which is better at surviving the winter (lower $d$). Neither type has an advantage over a full environmental cycle, and they stably coexist. This is due to a combination of the storage and bounded density effects (recall that stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).

The classic lottery model (Eq. 1) fails to give co-existence for these parameters because expansion to carrying capacity occurs immediately at the start of the summer (Fig. \ref{fig:fluctuatingselection}d-f). As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than required when we properly account for the growth in the abundance of each type using Eq. \eqref{eq:master} (keeping the other parameters the same; Fig.~\ref{fig:fluctuatingselection}g-i). Previous models of the promotion of genetic variation via the storage effect  \citep{ellner_1994} similarly assume that the total number of offspring per iteration is constant, and would produce a similar error.

@@

\begin{figure}
\centering
\includegraphics[scale=0.7]{fluctuatingselection.pdf}
\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment requires a much greater $b$ advantage in the classic lottery model compared to our density-dependent extension of it when population density is seasonally cyclical. (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer). The $b$-specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$-specialist ($b=0.217$, $d=0.1$) (blue). (b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$-specialist does so at a lower rate. Total height (blue+black) is population density $N/T$. (c) Summer favors the $b$ specialist, winter the $d$-specialist, and they stably coexist. (d-f) Same as (a-c) for the classic lottery model; the types no longer coexist. (g-i) Same as (d-f) where now $b = 0.0421$ for the $d$ specialist and the types coexist. For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed dispersal.} 
\end{figure}

@@

It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme, where $r=b-d$ is the maximal, low-density growth rate \citep{pianka_1972}. Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}. Contrary to predictions of an $r$/$K$ trade-off, empirical studies have shown that maximal growth rate at low density and the high density at which saturation occurs (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}. From the perspective of our model, this positive correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.

@@

$K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given $r$, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements. To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness, since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density per territory $L$. The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different genotypes. We leave these complications for future work.

Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations that can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}. Moreover, our formulation is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g. \citealt{tilman_2007}). However, section ``Primary strategies and Grime's triangle'' suggests that, apart from the obvious irrelevance of territorial contests at low density, selection alone has no preference for fecundity, competitive ability or longevity. One or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation.

Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$-specialists), may also be strongly $d$-selected, which while unintuitive, is consistent with our findings. An effective way to reduce $d$ in the face of unavoidable physical destruction is to shorten the time to reproductive maturity --- short life cycles are a characteristically ruderal trait. Moreover, a recent hierarchical cluster analysis of coral traits did find a distinct ``ruderal'' cluster, but high fecundity was not its distinguishing feature. Rather, ruderals used brood- (as opposed to broadcast-) spawning, which could plausibly be a mechanism for improving propagule survivorship in disturbed environments \citep{darling_2012}.

One potential limitation of our model as a general-purpose model of density-dependent selection is its restriction to interference competition between juveniles for durable resources (lottery recruitment to adulthood), analogous to the ubiquitous assumption of viability selection in population genetics \citep[p. 45]{ewens_2004}. In some respects this is the complement of consumable resource competition models, which restrict their attention to indirect exploitation competition, typically without age structure \citep{tilman_1982}. In the particular case that consumable resources are spatially localized (e.g. due to restricted movement through soils), resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition). The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked. It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar genotypes that use identical resources rather than the stable co-existence of widely differing species with different resource preferences \citep{ram_2016}. We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources. More work is needed connecting resource competition models to the density-dependent selection literature, since most of the former has to date been focused on narrower issues of the role of competition at low resource availability and in the absence of direct interactions between organisms at the same trophic level \citep{aerts_1999,davis_1998,tilman_2007}.

@@

In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference. It seems that Eq. \eqref{eq:master} or something similar (and hopefully more analytically tractable) is unavoidable for the analysis of time-course genetic data because, fundamentally, selective births and deaths affect both abundances and frequencies, not one or the other in isolation. Moreover, some aspects of allele frequency change are intrinsically density-dependent. In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model''). This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist. This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}. It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}). In cyclical environments, polymorphisms can be stabilized by the bounded density effect, which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}. We leave the details of how our model might be applied to inference problems, including the crucial issue of its genetic drift predictions (providing a null model for neutral sites), for future work.

@@

In our representation of Grime's ideal environment, $d\ll 1$, $b_i/d_i \gg 1$ and $L\gg 1$, and so Eq. \eqref{eq:invad} becomes 
\begin{equation}
\Delta n_j/n_j \approx \frac{b_j}{c_j/c_i + L-1}\frac{c_j}{c_i}-d_j\approx \frac{b_j}{L}\frac{c_j}{c_i}-d_j, \label{eq:highLinvad}
\end{equation}
where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i$ (ignoring mutations with very large effect sizes; incidentally, Eq. \eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this assumption is unrealistic for a rare mutant). Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different absolute magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium. Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy. The same argument applies for the high-$L$ interpretation of the stress regime.

@@

The effects of improving $b$ and $d$ are now no longer identical for given values of $d_{ie}$ and $d_{ig}$. A $b$-mutation $b_j=(1+\epsilon)b_i$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i \approx \epsilon$, whereas a $d$-mutation $d_{jg}=(1-\epsilon) d_{ig}$ gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$. Since overall mortality is high ($d_i\approx 1$), $d_{g}$ can be small compared to $d_{e}$ and yet still make a greater contribution to invasion fitness than $b$. The case where $d_{ig}$ is so small that $d_{ig}\ll 1-d_i$ (i.e. $d_{ig}$ is an order of magnitude smaller than $1-d_i$ which itself an order of magnitude smaller than $d_{ie}$) effectively represents a hard constraint that $d$ cannot evolve. We do not deny this possibility, but this brings us back to issues of mutational bias and constraint (the first two factors controlling the direction of evolution in section ``Primary strategies and Grime's triangle''), which we do not address in this manuscript.
